---
title: "Homework 2"
subtitle:  "Joint, Marginal and Conditional"
graphics: yes
output: pdf_document
header-includes:
    - \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* * * 
### Instructions

Please answer the following questions in the order in which they are posed. Add a few empty lines below each and write your answers there. Focus on answering in complete sentences. You will also need scratch paper/pen to work out the answers before typing it.

For help with formatting documents in RMarkdown, please consult [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/). Another option is to search using Google. 

* * *

### Exercises 

1. (Joint PMF) An urn contains four red chips, three white chips,
and two blue chips. A random sample of size 3 is drawn
without replacement. Let $X$ denote the number of white
chips in the sample and $Y$ the number of blue chips. Write
a formula for the joint PMF of $X$ and $Y$. Be sure to explain the components of your formula and why it makes sense. 

   \emph{Hint: You will first want to make a table on scratch paper to understand the possible values for $x$ and $y$ and their corresponding probabilities. Then the formula for the PMF will be evident}

Suppose that there are 4 red chips, 3 white chips, and 2 blue chips in which 3 chips are drawn at random without replacement. Let $X$ denote the number of white chips drawn and $Y$ denote the number of blue chips drawn. The Joint PMF of $X$ and $Y$ is shown in the table below.
```{r Joint PMF}
col1 <- c(1/21, 1/7, 1/21)
col2 <- c(3/14, 2/7, 1/28)
col3 <- c(1/7, 1/14, 0)
col4 <- c(1/84, 0, 0)
(joint_PMF <- data.frame("Y" = c(0,1,2), "0" = col1, "1" = col2, "2" = col3, "3" = col4))
```

The formula for the joint PMF of $X$ and $Y$ is,
$$f(x,y) = \frac{ {4 \choose {4-x-y}}{3 \choose x}{2 \choose y} }{ {9 \choose 3} } \text{ for } x = 0,1,2,3 \text{ and } y=0,1,2 $$
To get this PMF equation, we have to keep in mind how many total chips there are and how many of each chip is drawn for each value of $X$ and $Y$. We know that there are only 3 white chips and 2 blue chips, hence $X$ and $Y$ can take values of $x = 0,1,2,3, y=0,1,2$. Thus we can describe the number of white marbles drawn as ${3 \choose x}$ and the number of blue marbles as ${2 \choose y}$ for each 3 draws without replacement. Then the remaining marbles will be red, which we can describe as ${4 \choose {4-x-y}}$ (because there are 4 total red marbles and $4-x-y$ marbles left to draw). Then we need to divide by the total number of marbles drawn, which we write as ${9 \choose 3}$ because there are 9 total chips and we are drawing 3 of them. Putting this all together we get the PMF formula.

2. (Multinomial) Let $U_1,U_2, . . . ,U_{1029}$ be independent
uniformly distributed random variables. Let $X_1$ equal the number of $U_i$ less than
.331, $X_2$ equal the number between .331 and .820, and $X_3$ equal the number greater than .820.


   a. Calculate, using, R the probability of observing $X_1 = 354, X_2 = 492, X_3 = 183$. Be sure to show your code. State the joint distribution before launching into calculations. 
   
Let $U_1,U_2, . . . ,U_{1029}$ be independent uniformly distributed random variables where $X_1$ equal the number of $U_i$ less than .331, $X_2$ equal the number between .331 and .820, and $X_3$ equal the number greater than .820. Thus we know that $U_i \sim Unif(0,1)$ and we also know that the PDF of a uniform random variable is $f(x) = \frac{1}{a-b}$ for $a \leq x <b$ on the interval $[a,b)$.

Then to find the probability that $X_1 = 354, X_2 = 492, X_3 = 183$, we first have to find the probabilities of landing in the ranges represented by $X_1, X_2,$ and $X_3$. Let $\pi_1, \pi_2, \pi_3$ represent these probabilities respectively. Since we know the PDF of $U_i$ is $f(u) = 1$ on the interval $0 \leq u < 1$. Then,
\begin{align*}
\pi_1 &= \int_{a}^{b} f(u) du \\
      &= \int_{0}^{0.331} 1 du \\
      &= 0.331 - 0 = 0.331 \\
\end{align*}
\begin{align*}
\pi_2 &= \int_{0.331}^{0.820} 1 du \\
      &= 0.820 - 0.331 = 0.489 \\
\end{align*}
\begin{align*}
\pi_3 &= \int_{0.820}^{1} 1 du \\
      &= 1 - 0.820 = 0.180 \\
\end{align*}

Now we can use a multinational distribution to find the probability of $X_1 = 354, X_2 = 492, X_3 = 183$. In order to do this we hae some assumptions.

1. The number of trials is fixed at 1029, where each trial is a draw from a $Unif \sim (0,1)$.

2. Probability of each trial is the same in each trial (the probabilities of landing in each ranges represented by $X_1, X_2,$ and $X_3$) that we just calculated.

3. There are 3 outcomes in each trial, the draw is in one of 3 ranges, 0 to 0.331, 0.332 to 0.820, or 0.882 to 1.

4. Finally we have to assume that each trial is independent. Meaning that the probability of one draw will not affect the outcome of another draw (draws from the $Unif \sim (0,1)$).

Hence we have $<X_1, X_2, X_3> \sim Multinational(n=1029,\pi=<0.331,0.489,0.180>$ and the PDF is $f(x_1,x_2,x_3) = \frac{1029}{x_1x_2x_3}(0.331)^{x_1}(0.489)^{x_2}(0.180)^{x_3}$. Then,

```{r multinom}
prob <- dmultinom(x = c(354, 492, 183), size = 1029, prob = c(0.331,0.489,0.180))
```
Thus the probability of observing $X_1 = 354, X_2 = 492, X_3 = 183$ is `r prob`.

   b. Calculate, using R as a calculator, the expected values and standard deviation of $X_2$? Be sure to show your code. State the marginal distribution before launching into calculations.

We know that the marginal distribution of $X_2$  is $Binom\sim(n,\pi)$, thus the expected values and standard deviation of $Binom\sim(n,\pi)$ are $E[X]=n\pi$ and $\sqrt {Var[X]} = \sqrt {n\pi(1-\pi)}$. Then,
```{r expected and sd}
x_2_expected <- 1029 * 0.489
x_2_sd <- sqrt(1029 * (0.489) * (1-0.489))
```
Thus we get $E[X_2] =$ `r x_2_expected` and $SD[X_2] =$ `r x_2_sd`.

   
3. (Marginal PMF) Let $X$ and $Y$ be discrete random variables with joint PMF 
$$f(x, y) = \left(\frac{\lambda}{2} \right)^{y} \frac{e^{-\lambda}}{x! \:(y-x)!}$$

   where $x$ and $y$ are (non-negative) integers and $0 \leq x \leq y$. That is, $x, y = 0, 1, 2, 3, \dots$ but with the constraint $0 \leq x \leq y$.

   Determine $f_2(y)$, the marginal distribution of $Y$. Is this a familiar distribution? Show the steps.
   
      \emph{Hint: you will need to use the Binomial theorem with a = 1 and b = 1 to perform the summation over $x$. }  
      
By definition of marginal distribution,
\begin{align*}
f_2(y) &= \sum_{x} f(x,y) \\
       &= \text{ since x takes on values of 0,1,2,3 and }  0 \leq x \leq y \\
       &= \sum_{x=0}^{y} (\frac{\lambda}{2})^y \frac{e^{-\lambda}}{x!(y-x)!} \\
       &= (\frac{\lambda}{2})^y e^{-\lambda} \sum_{x=0}^{y} \frac{1}{x!(y-x)!} \\
       &= \text{ then we can multiply by y! inside the sumation and } \frac{1}{y!} \text{ outside the sumation} \\
       &= \frac{1}{y!} (\frac{\lambda}{2})^y e^{-\lambda} \sum_{x=0}^{y} \frac{y!}{x!(y-x)!} \\
       &= \frac{1}{y!} (\frac{\lambda}{2})^y e^{-\lambda} \sum_{x=0}^{y} {y \choose x} \\
       &= \text{ note that the summation term can be replaced by } (1+1)^y \\
       &= \text{ because by the binomial theorem, } (1+1)^y = \sum_{x=0}^{y} {y \choose x} 1^{x-y} 1^{x} \\
       &= \frac{1}{y!} (\frac{\lambda}{2})^y e^{-\lambda}(1+1)^y \\
       &= \frac{e^{-\lambda}\lambda^y}{y!}
\end{align*}

We can see that the marginal distribution of $Y$ is $\frac{e^{-\lambda}\lambda^y}{y!}$. Notice that $f_2(y)$ is the same as the PMF of a Poisson distribution where $\lambda$ represents the number of occurrences of event Y over a fixed time period.

4. (Hierarchical model) Suppose a player is equally likely to have 4, 5 or 6 at-bats (opportunities to bat) in a baseball game. If $X$ is the number of opportunities to bat, then we are assuming that
$$f_1(x) = P(X = x) = \frac{1}{3}, \ \ \ x = 4, 5,6.$$
Suppose $Y$, the number of hits, is a Binomial random variable with size $X = x$ and probability of success $\pi=0.3$. That is 
$$f(y|x) = P(Y = y | X = x) = Binom(x, 0.3).$$
 
a. Fill in the numbers for the joint PMF, $f(x,y) = P(X = x, Y = y)$ in the cells indicated by (i) -- (v). Also fill in the number for the marginal PMF $f_2(y)$ in the cell indicated by (vi). Show work below the table so we know you are not just guessing.

\begin{table}[h]
\centering
\begin{tabular}{||c|c|c|c|c|c|c|c||} \hline
& \multicolumn{7}{c||}{$y$} \\ \hline
$x$ & 0 & 1 & 2 & 3 & 4 & 5 & 6  \\ \hline
4 & \hspace*{0.2in} & \hspace*{0.2in} & (i) & \hspace*{0.2in}& (ii) & (iii) & \\ \hline \hline
5 & & & (iv) & & & &  \\ \hline  \hline
6 & & & (v) & & & &   \\ \hline \hline
$f_2(y)$ & & & (vi) & & & & \\ \hline
\end{tabular}
\end{table}

First we need to find the formula for the joint PMF of $X$ and $Y$. By definition of conditional distributions, we know that,
\begin{align*}
f(y|x) &= \frac{f(x,y)}{f_1(x)} \\
       &= \text{ since } f_1(x) \text{ is given to be } 1/3 \\
       &= \frac{f(x,y)}{1/3} \\
       &= 3f(x,y)\\
f(x,y) &= \frac{f(y|x)}{3} \\
       &= f(Y|X) \text{ was given to be a binomal random variable with size $X = x$ and probability of success $\pi=0.3$} \\
       &= \frac{Binom(X=x, \pi=0.3)}{3} \\
       &= \text{ by definition of binomial random variable} \\
       &= \frac{{x \choose y}0.3^y(1-0.3)^{x-y}}{3} \\
\end{align*}
Now using the joint PMF formula we can find specific probabilities. 
```{r specific probs}
(i <- choose(4, 2) * 0.3^2 * (1-0.3)^(4-2)/3)
(ii <- choose(4, 4) * 0.3^4 * (1-0.3)^(4-4)/3)
(iii <- choose(4, 5) * 0.3^5 * (1-0.3)^(4-5)/3)
(iv <- choose(5, 2) * 0.3^2 * (1-0.3)^(5-2)/3)
(v <- choose(6, 2) * 0.3^2 * (1-0.3)^(6-2)/3)
(vi <- i + iv + v)
```
Then filling in the table, 
\begin{table}[h]
\centering
\begin{tabular}{||c|c|c|c|c|c|c|c||} \hline
& \multicolumn{7}{c||}{$y$} \\ \hline
$x$ & 0 & 1 & 2 & 3 & 4 & 5 & 6  \\ \hline
4 & \hspace*{0.2in} & \hspace*{0.2in} & (`r i`) & \hspace*{0.2in}& (`r ii`) & (`r iii`) & \\ \hline \hline
5 & & & (`r iv`) & & & &  \\ \hline  \hline
6 & & & (`r v`) & & & &   \\ \hline \hline
$f_2(y)$ & & & (`r vi`) & & & & \\ \hline
\end{tabular}
\end{table}

b. Write the conditional distribution of $X$ given $Y=2$. That is $$f(x|y=2) = P(X = x | Y = 2).$$ Again, be sure to state any formulas you are plugging into so we know you are not guessing.

By definition of conditional distributions, we know that,
\begin{align*}
f(x|y) &= \frac{f(x,y)}{f_2(y)} \\
f(x|y=2) &= \frac{f(x,2)}{f_2(2)} \\
         &= \text{ from part a we know } f_2(2) = `r vi` \\
         &= \frac{f(x,2)}{`r vi`} \\
         &= \text{ using our joint PMF formula found in part a} \\
         &= \frac{{x \choose 2}0.3^2(1-0.3)^{x-2}}{3}\frac{1}{`r vi`} \\
         &= {x \choose 2} \cdot (0.7)^x \cdot (0.7)^{-2} \cdot (0.3)^2 \cdot \frac{1}{3 \cdot `r iv`} \\
         &= \text{ which is about} \\
         &= {x \choose 2} \cdot (0.7)^x \cdot 0.205 \\
\end{align*}

Hence $f(x|y=2) = {x \choose 2} \cdot (0.7)^x \cdot 0.205$ for $x=4,5,6$.
